Introduction to *cellcount*

### 2022-09-20 cellcount_intro.Rmd
Updates to markdown - NOT COMPLETED

---
title: "Introduction to *cellcount*"
author: "*T.E. Harman, B. Boyd, K. Pokrzywinski*"
date: "11/2/2021"
abstract: >
  The *cellcount* package provides wrapper functions for  [*EBImage*](http://bioconductor.org/packages/devel/bioc/vignettes/EBImage/inst/doc/EBImage-introduction.html) to simplfy quantifying bacteria and cyanobacteria via flourescence microscopy. Cyanobacteria enumeration is an important tool to incorporate data results into mathematical models to assess bloom dynamics. However, standard manual cyanobacteria enumeration is considered time-consuming and can impact overall random error due to different inaccuracies from researcher to researcher. Here, we outline a new open-source tool, *cellcount*, a package designed for the computing language R to assist with cyanobacteria enumeration. Fluorescent microscopy images are analyzed by this package to generate cyanobacteria counts to assess overall cell density from samples.
output: html_document
---

@@

The following packages are required. Please install and bring them into the R
environment: 
```{r, message=FALSE}
if (!require("EBImage")) {
  BiocManager::install("EBImage")
}
library(EBImage)
library(tiff)
library(pixmap)
library(raster)
```

@@

You'll need to generate a couple of directories; one for saving the unique
color label images and another for CSV files. In this example, we are using 
images of vegetative *Alexandrium catenella* cells. Please note that your 
directory addresses will be different. We recommend using a dedicated directory 
for *cellcount* results:

```{r}
savdir <- ("E:/cellcount_data/CSV_data/")
image_savdir <- ("E:/cellcount_data/Convert_Images/Convert_Alexandrium10X/")
```

Next, you'll need to read in the images you are interested in processing using
the following code. Once again, your directories will be different and must be
updated to the correct file address:

```{r}
images <- list.files("E:/cellcount_data/Alexandrium_10X/", pattern = "tif", full.name = T)
images_names <- list.files("E:/cellcount_data/Alexandrium_10X/", pattern = "tif", full.name = F)
```

@@

```{r}
imgNames <- paste0("Alexandrium_", images_names)
```

Now we will "read" in the images that we are interested in analyzing. The
following code outlines this step using the "readTIFF" function from the 
*tiff* package:

```{r, warning=FALSE}
read_images <- lapply(images, readTIFF)
names(images) <- imgNames
```

@@

```{r}
grey_images <- lapply(read_images, greyscale, contrast = 2)
display(grey_images[[1]])
```

@@

Before we analyze our images in bulk, set aside some code space to visually 
analyze your images results. This could include adjusting the threshold
from the "mapped" function, the contrast from the "greyscale" function 
(if used), or could involve making adjustments to the "image_convert" function,
such as the width and height of the rectangular window or the pixel area
threshold (i.e., removes cells or other objects under a certain pixel area).
Observe your initial results and make adjustments where necessary:

```{r}
imagesConverted <- image_convert(imagesMapped[[1]], w = 15, h = 25, offset = 0.01, areathresh = 150)
final_img <- countImages(imagesConverted, normalize = T, removeEdgeCells = T)
display(final_img)
```

@@

Once you have made necessary adjustments, use the following for-loop to
analyze your images in bulk, one at a time. The for-loop uses the 
"image_convert" function to produce unique color labels on cells and produce an
image showcasing these labels using the "countImages" function. Finally, 
"countCells" is used to count the number of unique color labels (i.e., cells).
We also have integrated code to add these count results to the data frame made
earlier, and also to save each unique color label image to the "image_savdir" 
directory:

```{r}
for (z in 1:length(images)) {
  imagesConverted <- image_convert(imagesMapped[[z]], w = 15, h = 25, offset = 0.01, areathresh = 150)
  final_img <- countImages(imagesConverted, normalize = T, removeEdgeCells = T)
  count <- countCells(imagesConverted)
  Cell.Count[nrow(Cell.Count) + 1, ] <- c(imgNames[[z]], count)
  # display single-cell images via image_covert
  analyzed_image <- paste0(sub(".tif", replacement = "", x = imgNames[z]), "_analyzed.tiff")
  writeImage(final_img, files = paste0(image_savdir, analyzed_image))
}
```

@@

@@

```{r}
cell.total <- as.numeric(Cell.Count$Cell_Count)
cell.total <- sum(cell.total)
Cell.Count[nrow(Cell.Count) + 1, ] <- c("Cell Total", cell.total)
```

@@

- Total counts (generated previously)
- Height of image field of view (μm)
- Width of image field of view (μm)
- Median cell diameter (μm)
- Total number of images analyzed
- Filtration area (mm2)
- Volume filtered per image (mL)
- Total sample volume (mL)

@@

```{r}
cell.den <- cell_density(cell.total, height = 203.2, width = 166.4, diameter = 2.5, images = 15, filtration.area = 491, volume = 1, total.volume = 100)
Cell.Count[nrow(Cell.Count) + 1, ] <- c("Cell Density", cell.den)
```

>>>>

---
title: "Introduction to *cellcount*"
author: "*T.E. Harman, M. Vandersea, D.E. Berthold, H.D. Laughinghouse, K. Pokrzywinski*"
date: "9/__/2022"
abstract: >
  The *cellcount* package provides wrapper functions for  [*EBImage*](http://bioconductor.org/packages/devel/bioc/vignettes/EBImage/inst/doc/EBImage-introduction.html) to simplfy quantifying bacteria and cyanobacteria via flourescence microscopy. Cyanobacteria enumeration is an important tool to incorporate data results into mathematical models to assess bloom dynamics. However, standard manual cyanobacteria enumeration is considered time-consuming and can impact overall random error due to different inaccuracies from researcher to researcher. Here, we outline a new open-source tool, *cellcount*, a package designed for the computing language R to assist with cyanobacteria enumeration. Fluorescent microscopy images are analyzed by this package to generate cyanobacteria counts to assess overall cell density from samples.
output: html_document
---

@@

The following packages are required. Please install and bring them into the R
environment: 
```{r, message=FALSE}
if (!require("EBImage")) {
  BiocManager::install("EBImage")
}
library(EBImage)
library(tiff)
library(pixmap)
library(raster)
library(beepr)
```

@@

You'll need to generate a couple of directories; one for saving the unique
color label images and another for CSV files. In this example, we are using 
images of *Microcystis* sp. cells. Please note that your 
directory addresses will be different. We recommend using a dedicated directory 
for *cellcount* results:

```{r}
savdir <- ("H:/CSV_data/")
image_savdir <- ("H:/Convert_Images/F192_Test/")
```

Next, you'll need to read in the images you are interested in processing using
the following code. Once again, your directories will be different and must be
updated to the correct file address. However, in this example we will be using
the provided test images from this package:

```{r}
images <- list.files("C:/Users/Tyler.Harman/Desktop/cellcount_work/cellcount_main/Microcystis_F192_40x/", pattern = "tif", full.name = T)
images_names <- list.files("C:/Users/Tyler.Harman/Desktop/cellcount_work/cellcount_main/Microcystis_F192_40x/", pattern = "tif", full.name = F)
```

@@

```{r}
file_name<-paste0("F192_")
```

Now we will "read" in the images that we are interested in analyzing. The
following code outlines this step using the "readTIFF" function from the 
*tiff* package. In addition, we'll need to re-orient the image as it's uploaded
into the environment. The "readTIFF" function uploads images in a portrait 
format - the "aperm" function reorients the uploaded image to a landscape format:

```{r, warning=FALSE}
imgNames <- paste0(file_name, images_names)
read_images <- lapply(images, readTIFF)
img_transposed <- lapply(read_images,aperm,c(2,1,3))
names(images) <- imgNames
```

@@

```{r}
grey_images <- lapply(img_transposed, greyscale, contrast = 1)
display(grey_images[[5]])
```

@@

Before we analyze our images in bulk, set aside some code space to visually 
analyze your images results. This could include adjusting the threshold
from the "mapped" function, the contrast from the "greyscale" function 
(if used), or could involve making adjustments to the "image_convert" function,
such as the width and height of the rectangular window, the pixel area
threshold (i.e., removes cells or other objects under a certain pixel area), 
or the tolerance and 'ext' variables. Observe your initial results as indicated
below and make adjustments where necessary:

```{r}
imagesConverted <- image_convert(imagesMapped[[5]], w = 10, h = 10, offset = 0.001, areathresh = 50, tolerance = 0.8, ext = 1)
final_img <- countImages(imagesConverted, normalize = T, removeEdgeCells = T)
display(final_img)
```

@@

Once you have made necessary adjustments, use the following for-loop to
analyze your images in bulk, one at a time. The for-loop uses the 
"image_convert" function to produce unique color labels on cells within
an image and to produce a final image showcasing these labels using the 
"countImages" function. Finally, "countCells" is used to count the number of 
unique color labels (i.e., cells). We also have integrated code to add these 
count results to the data frame made earlier, and also to save each unique 
color label image to the "image_savdir" directory:

```{r}
for (z in 1:length(images)) {
  imagesConverted <- image_convert(imagesMapped[[z]], w = 10, h = 10, offset = 0.001, areathresh = 50, tolerance = 0.8, ext = 1)
  final_img <- countImages(imagesConverted, normalize = T, removeEdgeCells = T)
  count <- countCells(imagesConverted)
  Cell.Count[nrow(Cell.Count) + 1, ] <- c(imgNames[[z]], count)
  analyzed_image <- paste0(sub(".tif", replacement = "", x = imgNames[z]), "_analyzed.tiff")
  writeImage(final_img, files = paste0(image_savdir, analyzed_image),compression=c("LZW"))
}
```

@@

Some images that were initially taken may be too poor of quality or contain
high background fluorescence, among other errors. The code below analyzes 
outliers from the produced data, within a range of two standard deviations 
above and below the median. The "outlier" function tags which images are
out of this range and dictates a warning within our data frame, which is shown 
below. This alerts the end-user to perhaps re-image a particular sample, or
to investigate the error further.

```{r}
Cell.Count$Cell_Count<-as.numeric(Cell.Count$Cell_Count)
Cell.Count$SD_range<-outlier()
Cell.Count
```

@@

```{r}
cell.total <- as.numeric(Cell.Count$Cell_Count)
cell.total <- sum(cell.total)
Cell.Count[nrow(Cell.Count) + 1, ] <- c("Cell Total", cell.total,"null")
```

*Important note* - *the "null" data produced in the code above (and others that*
*follow) fills in the blanks from the previous data produced by the* 
*"outlier" function.*

@@

- Total counts (generated previously)
- Height of image field of view (μm)
- Width of image field of view (μm)
- Median cell diameter (μm)
- Total number of images analyzed
- Filtration area (mm^2)
- Volume filtered per image (mL)
- Total sample volume (mL)

@@

```{r}
cell.den <- cell_density(cell.total, height = 203.2, width = 166.4, diameter = 2.5, images = 15,
                         filtration.area = 491, volume = 1, total.volume = 100)
Cell.Count[nrow(Cell.Count) + 1, ] <- c("Cell Density", cell.den,"null")
```

We can also calculate the average number of cells per image, as well as cells
per mL using the following code:

```{r}
cell.av<-(cell.total/10)
Cell.Count[nrow(Cell.Count) + 1, ] <- c("Imaging Average",cell.av,"null")

cell.mL<-(cell.den/2) #change this via total volume
Cell.Count[nrow(Cell.Count) + 1, ] <- c("Total Volume Cell per mL",cell.mL,"null")
```

We can now view our data frame to double check for any calculation errors:
